Optogenetic Maxwell Demon to Exploit Intrinsic Noise and Control Cell Differentiation Despite Time Delays and Extrinsic Variability

### 2023-01-31 paper.tex
Update introduction - needs some citations but is getting close.

In \cite{May2021}, it was surprisingly found that, when noise, auto-regulation, and noise-exploiting feedback control were all present at once, the system could be controlled to bring two similar cells to two different arbitrary points simultaneously (Figure \ref{cartoons}A), but when any one of these components were missing the system could not break its symmetry. More specifically, this required that the symmetry between the protein probability distributions of each cell would not be not the same (or asymmetric). Three effective controllers were found which could do this, a Fully Aware Feedback Controller (FAC) that observed the protein counts of all observed cells, a Partially Aware Feedback Controller (PAC) that observed the protein counts of only one cell, and a probabilistic Model Predictive Feedback Controller (pMPC) that observed only the primary cell and integrated a probabilistic model to represent the non-stationary probability distribution of the secondary cells in the same environment.

The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states. For a two cell, single-input-single output (SIMO) control problem, the FAC performed the best (figure\ \ref{cartoons}C(top)), and the PAC (figure\ \ref{cartoons}C(bottom)) 2nd, and pMPC controllers 3rd. In all cases, the controller's knowledge and explicit treatment of intrinsic noise was critical to improve the control performance, break symmetry, and enable the SIMO control of all cells.

In this paper, the control problem is extended to examine how the FAC and PAC controllers are affected by time delay, the system granularity, and parameter uncertainty. In the `Methods' section, we briefly set up a stochastic chemical master equation model for multiple individual cells in the same spatially homogeneous environment, and we formulate the effects of FAC and PAC controllers on the dynamics of cells within that environment. Next, in the `Results' section, we examine the effects of volume changes, time delays, and parametric uncertainties on the FAC and PAC control performance. Finally, in `Conclusions' we summarize our findings and discuss potential implications that these controllers may have on future systems and synthetic biology investigations.

>>>>

Szymanska et al.~\cite{Szymanska2015} examined a theoretical system inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, where two gene products repress one another, but where degradation of one protein was under the control of UV radiation via the XXX pathway. Their analysis showed that control was possible despite certain parameter errors and time delays due to maturation of fluorescent proteins and limited observation of the regulatory proteins. 
To explore this concept in the context of more realistic parameters, May et al.~\cite{May2021} fit a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase. An extended model showed that addition of a positive auto-regulation module (see Figure XXX) could help to maintain an elevated expression phenotype in the presence of UV excitation. Using discrete stochastic models based on direct solutions to the chemical master equation, it was shown that   the combination of biochemical noise, nonlinear auto-regulation, and a {\em single} optogenetic feedback could allow two genetically identical cells with arbitrary initial conditions to be controlled simultaneously to reach different specified phenotypes (Figure \ref{cartoons}A). Moreover, effective control was possible not only for a Fully Aware Feedback Controller (FAC) that observed the protein counts of all observed cells, but control could be achieved almost as well for a Partially Aware Feedback Controller (PAC) that observed the protein counts of only one cell and used a probabilistic model to estimate the expression for unobserved cells. Moreover, it was again seen that control remained effective even in the circumstance where the model used to design the control law was an inexact simplification of the process dynamics to which the controller was later applied.  
%The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states. For a two cell, single-input-single output (SIMO) control problem, the FAC performed the best (figure\ \ref{cartoons}C(top)), and the PAC (figure\ \ref{cartoons}C(bottom)) 2nd, and pMPC controllers 3rd. In all cases, the controller's knowledge and explicit treatment of intrinsic noise was critical to improve the control performance, break symmetry, and enable the SIMO control of all cells.

In this paper, the control problem is extended to examine how the fully- and partially-aware controllers (FAC and PAC, respectively) are affected by additional model uncertainties, including coarse grained approximations to the system dynamics, uncertainties or errors in system parameters, and delays in the biochemical sensors or optogenetic controllers through which the external feedback is implemented. First, in the next section, we briefly describe our methods to set up a stochastic chemical master equation model for multiple individual cells in the same spatially homogeneous environment; we formulate the FAC and PAC controllers on the dynamics of cells within that environment; and we explain how the control law is optimized to achieve desired performance objectives. Next, in the `Results' section, we examine how the granularity of the model formulation, the inaccuracy of parametric assumptions, and the inclusion of time delays affect the FAC and PAC control performance. Finally, in `Conclusions' we summarize our findings and discuss the implications that these controllers, and their robustness to modeling errors, may have on future systems in synthetic biology.
